Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-10-20 pot/LaTeX/skewnormal.tex
time series updates

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
We assume the data can be modeled using a skew-$t$ distribution.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha \sigma_t z_t + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N_{(0, \infty)}(0, 1)$ are a random effect from a half-normal distribution (see appendix A.4), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

@@

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $\chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

\subsection{Accounting for temporal dependence}
We account for temporal dependence through a AR(1) time series on the spatial knots and $z_t(\bs)$ terms.
These are modeled as
\begin{align}
  \bw_{t, k} | \bw_{t-1, k} &\sim N_2(\phi_w \bw_{t-1, k}, (1 - \phi_w^2) ) \\
  z_{t, k} | z_{t-1, k} &\sim N(\phi_z z_{t-1, k}, (1 - \phi_z^2))
\end{align}
where $|\phi_w|, |\phi_z| < 1$.

\subsection{Hierarchical model}\label{s:hier}
Conditioned on $z_{tk}(\bs) \iid $ $N(0, \sigma^2_{tk})$, $\sigma^2_{tk}(\bs) \sim IG(a, b)$, and $P_{tk}$, the marginal distributions are skew-$t$ and the joint distribution within a partition is multivariate skew-$t$.
However, we do not fix the partitions, they are treated as unknown and updated in the MCMC.
We model this with a Bayesian hierarchical model as follows.
Let $\bw_{t1}, \ldots, \bw_{tK}$ be a set of daily spatial knots in a spatial domain of interest, $\calD$, so that
\begin{align*}
 P_{tk} = \{\bs : k = \argmin_\ell || \bs_t - \bw_{t\ell} ||\}.
\end{align*}
Because the spatial domain is bounded, we do a transformation on the knot locations to $\calR^2$.
Let $\Phi_2$ represent a bivariate standard normal density function, and let $\bw_k = \min(\bs) + \text{range}(\bs) * \Phi_2^{-1}(\bw^*_k)$.
Then $\bw^*_k \in \calR^2$.
Then
\begin{align}
   Y_t(\bs) \mid z_{t}(\bs), \sigma^2(\bs), P_{tk}, \alpha, \beta, \Theta &= X_t(\bs) \beta + \alpha |z_t(\bs)| + \sigma_t(\bs) v_t(\bs) \label{eq:hier}\\
   z_t(\bs) &= z_{tk} \text{ if } \bs \in P_{tk}\\
   \sigma^2_{t}(\bs) &= \sigma^2_{tk} \text{ if } \bs \in P_{tk}\\
   \alpha &\sim N(0, 10)\\
   v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma)\\
   z_{tk} \mid z_{t-1, k}, \sigma^2_{tk} &\sim N(\phi_z z_{t-1, k}, \sigma^2_{tk} (1 - \phi_z^2))\\
   \sigma^2_{tk} &\iid IG(\alpha, \beta)\\
   \bw^*_{tk} \mid \bw^*_{t-1, k} &\sim N_2(\phi_w \bw^*_{t-1, k}, (1 - \phi_z^2))
\end{align}
where $\Theta = \{\rho, \nu, \gamma\}$; \mbox{$k = \argmin_\ell ||\bs - \bw_\ell ||$}; and $\Sigma$ is a \Matern covariance matrix with variance one, spatial range $\rho$, smoothness $\nu$, and the proportion of variance accounted for by the spatial variation is $\gamma$.

@@

\subsection{Cross validation}\label{s:modelselect}
Models were compared using cross validation with 100 sites used as training sites and 30 sites witheld for testing.
The model was fit using the training set, and predictions were generated at the testing site locations.
Because one of the primary goals of this model is to predict extreme events, we quantify use Brier scores and quantile scores to select the model that best fits the data \citep{Gneiting2007}.
The Brier score for predicting exceedance of a threshold $c$ is given by $[e(c) - P(c)]^2$ where $e(c) = I[y>c]$ is an indicator function indicating that a test set value, $y$, has exceeded the threshold, $c$, and $P(c)$ is the predicted probability of exceeding $c$.
The quantile score for the $\tau$th quantile is $2\{ I[y < \widehat{q}(\tau)] - \tau\} (\widehat{q}(\tau) - y)$ where $y$ is a AQS test set value and $\widehat{q}(\tau)$ is the estimated $\tau$th quantile.
For both the Brier score and the quantile score, a lower score indicates a better fit.
These scores were averaged over all sites and days to obtain a single quantile score for each dataset.

@@

\subsection*{Spatial random effects}
If there is temporal dependence amongst the observations, then we update $z_tk$, using a Metropolis Hastings update.
First, we generate a candidate using a random walk Gaussian candidate distribution
\begin{align*}
  z_{tk}^{(c)} \sim \text{N}(z_{tk}^{(r - 1)}, s^2)
\end{align*}
where $z_{tk}^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
The acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[Y_t(\bs) | z_{tk}^{(c)}, \ldots] }{ l[Y_t(\bs) | z_{tk}^{(r - 1)}]} \right\} \times \left\{ \frac{ p[ z_{tk}^{(c)} ] }{ p[ z_{tk}^{(r - 1)}]}\right\}.
\end{align*}
The candidate is accepted with probability $\min\{R, 1\}$

>>>>

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
We assume the data can be modeled using a skew-$t$ distribution.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha |z_t| + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.4), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

@@

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Kim2005} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha |z_t(\bs)| + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $\chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

\subsection{Accounting for temporal dependence} \label{s:temporal}
In a standard block-maxima approach, the temporal dependence is assumed to be negligible because blocks (e.g. yearly maxima) are taken to be large enough to assume independence.
However, when using daily meausrements, the assumption of temporal independence is no longer appropriate.
We account for temporal dependence through an AR(1) time series on the spatial knots, $z_t(\bs)$ terms, and $\sigma^2_t(\bs)$ terms.
We first, transform the spatial knots from $\calD$ to $\calR^2$ as follows.
Let
\begin{align*}
  \bw^*_k = \Phi^{-1}_2\left[ \frac{ (\bw_k - \min(\bs))}{ \text{range}(\bs) } \right].
\end{align*}
where $\Phi_2$ represents a bivariate standard normal density function.
Then $\bw^*_k \in \calR^2$.
We use a copula on the $\sigma^2_t(\bs)$ to ensure that the marginal distributions of $\sigma^2_t(\bs)$ are inverse gamma.
Let
\begin{align*}
  \sigma^{2*}_t(\bs) =\Phi^{-1}\left\{ \text{IG}[\sigma^2_t(\bs)] \right\}
\end{align*}
where $\Phi$ is a univariate standard normal density function, and IG is the distribution function for an IG$(a, b)$ random variable.
Then the time series is modeled as
\begin{align}
  \bw^*_{t, k} | \bw^*_{t-1, k} &\sim N_2\left[\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2) \right] \\
  z_{t, k} | z_{t-1, k} &\sim N \left[\phi_z z_{t-1, k}, \sigma^2_{t,k} (1 - \phi_z^2)\right] \\
  \sigma^{2*}_{t, k} | \sigma^{2*}_{t-1, k} &\sim N \left[\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2) \right]
\end{align}
where $|\phi_w| < 1$, $0 < \phi_z < 1$, and $|\phi_\sigma| < 1$.

\subsection{Hierarchical model}\label{s:hier}
Conditioned on $z_{t,k}(\bs) \iid $ $N(0, \sigma^2_{t,k})$, $\sigma^2_{t,k}(\bs) \sim IG(a, b)$, and $P_{t,k}$, the marginal distributions are skew-$t$ and the joint distribution within a partition is multivariate skew-$t$.
However, we do not fix the partitions, they are treated as unknown and updated in the MCMC.
We model this with a Bayesian hierarchical model as follows.
Let $\bw_{t,1}, \ldots, \bw_{t,K}$ be a set of daily spatial knots in a spatial domain of interest, $\calD$, so that
\begin{align*}
 P_{t,k} = \{\bs : k = \argmin_\ell || \bs_t - \bw_{t\ell} ||\}.
\end{align*}

Then
\begin{align}
   Y_t(\bs) \mid z_{t}(\bs), \sigma^2(\bs), P_{t,k}, \alpha, \beta, \Theta &= X_t(\bs) \beta + \alpha |z_t(\bs)| + \sigma_t(\bs) v_t(\bs) \label{eq:hier}\\
   z_t(\bs) &= z_{t,k} \text{ if } \bs \in P_{t,k}\\
   \sigma^2_{t}(\bs) &= \sigma^2_{t,k} \text{ if } \bs \in P_{t,k}\\
   \alpha &\sim N(0, 10)\\
   v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma)\\
   z_{t,k} \mid z_{t-1, k}, \sigma^2_{tk} &\sim N(\phi_z z_{t-1, k}, \sigma^2_{t,k} (1 - \phi_z^2))\\
   \sigma^{2*}_{t,k} \mid \sigma^{2*}_{t-1, k} &\sim N(\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2))\\
   \bw^*_{t,k} \mid \bw^*_{t-1, k} &\sim N_2(\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2))
\end{align}
where $\Theta = \{\rho, \nu, \gamma\}$; \mbox{$k = \argmin_\ell ||\bs - \bw_\ell ||$}; and $\Sigma$ is a \Matern covariance matrix with variance one, spatial range $\rho$, smoothness $\nu$, and the proportion of variance accounted for by the spatial variation is $\gamma$.

@@

\subsection{Cross validation}\label{s:modelselect}
Models were compared using cross validation with 100 sites used as training sites and 30 sites witheld for testing.
The model was fit using the training set, and predictions were generated at the testing site locations.
Because one of the primary goals of this model is to predict extreme events, we quantify use Brier scores and quantile scores to select the model that best fits the data \citep{Gneiting2007}.
The Brier score for predicting exceedance of a threshold $c$ is given by $[e(c) - P(c)]^2$ where $e(c) = I[y>c]$ is an indicator function indicating that a test set value, $y$, has exceeded the threshold, $c$, and $P(c)$ is the predicted probability of exceeding $c$.
The quantile score for the $\tau$th quantile is $2\{ I[y < \widehat{q}(\tau)] - \tau\} (\widehat{q}(\tau) - y)$ where $y$ is a test set value and $\widehat{q}(\tau)$ is the estimated $\tau$th quantile.
For both the Brier score and the quantile score, a lower score indicates a better fit.
These scores were averaged over all sites and days to obtain a single quantile score for each dataset.

@@

\subsection*{Spatial random effects}
If there is temporal dependence amongst the observations, then we update $z_{tk}$, using a Metropolis Hastings update.
First, we generate a candidate using a random walk Gaussian candidate distribution
\begin{align*}
  z_{tk}^{(c)} \sim \text{N}(z_{tk}^{(r - 1)}, s^2)
\end{align*}
where $z_{tk}^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
The acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[Y_t(\bs) | z_{tk}^{(c)}, \ldots] }{ l[Y_t(\bs) | z_{tk}^{(r - 1)}]} \right\} \times \left\{ \frac{ p[ z_{tk}^{(c)} ] }{ p[ z_{tk}^{(r - 1)}]}\right\}.
\end{align*}
The candidate is accepted with probability $\min\{R, 1\}$
